---
title: "HW19: pca_taxon_stats"
author: "Bonnie Hurwitz"
date: "`r Sys.Date()`"
output:
  word_document:
    toc: yes
    toc_depth: '3'
  pdf_document:
    toc: yes
    toc_depth: '3'
  html_document:
    theme: paper
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: yes
subtitle: Exploring PCA and Taxon Stats
editor_options:
  markdown:
    wrap: sentence
---

# Setup

```{r setup, include=FALSE}
# Change your NETID
knitr::opts_chunk$set(echo = TRUE, fig.width = 6, fig.height = 4, dpi = 120)
knitr::opts_knit$set(root.dir = '/xdisk/bhurwitz/bh_class/YOUR_NETID/assignments/19_pca_taxon_stats')
```

```{bash, engine.opts='-l'}
# Let's make it so R can find the fonts we need
export FONTCONFIG_PATH=/etc/fonts
```

```{r}
dyn.load("/opt/ohpc/pub/apps/libpng/1.6.37/lib/libpng16.so.16")
```

```{r}
getwd()
```

```{r message=FALSE}
library(seriation)
library(dplyr)
library(purrr)
library(ggplot2)
library(phyloseq)
library(microViz)
library(shiny)
```

## Importing and fixing taxa your FMT dataset

To get started, we need to import our example.biom file into R and work with it using the phyloseq package

```{r}
# replace your NETID and SETID here, for example set21.biom
fmt <- import_biom("/xdisk/bhurwitz/bh_class/YOUR_NETID/assignments/12_phyloseq/YOUR_SETID.biom")
```

```{r}
tax_table(fmt) %>% head()
```

### Fixing tax_table

Let's format the dataset.

```{r}
fmt@tax_table@.Data <- substring(fmt@tax_table@.Data, 4)
colnames(fmt@tax_table@.Data)<- c("kingdom", "phylum", "class", "order", "family", "genus", "species")
fmt <- tax_fix(fmt, verbose = FALSE)
```

# Beta diversity (part 4)

## PCA

Principal Components Analysis.

Quite similar to Principal Co-ordinates Analysis.

In fact, PCA produces equivalent results to PCoA with euclidean distances.
So let's perform the CLR-transform first and check PCA and euclidean PCoA are the same.

```{r}
fmt %>%
  tax_filter(min_prevalence = 2.5 / 100, verbose = FALSE) %>%
  tax_transform(rank = "family", trans = "clr", zero_replace = "halfmin") %>%
  dist_calc(dist = "euclidean") %>%
  ord_calc(method = "PCoA") %>%
  ord_plot(alpha = 0.6, size = 2, color = "type") +
  theme_classic(12) +
  coord_fixed(0.7) +
  scale_y_reverse()
```

```{r}
fmt %>%
  tax_filter(min_prevalence = 2.5 / 100, verbose = FALSE) %>%
  tax_transform(rank = "family", trans = "clr", zero_replace = "halfmin") %>%
  ord_calc(method = "PCA") %>%
  ord_plot(alpha = 0.6, size = 2, color = "type") +
  theme_classic(12) +
  coord_fixed(0.7)
```

So why is PCA interesting for us?
Because the Principal components are built directly from a (linear) combination of the original features.

That means we know how much each taxon contributes to each PC axis, and we can plot this information (loadings) as arrows, alongside the sample points.

```{r}
pca <- fmt %>%
  tax_filter(min_prevalence = 2.5 / 100, verbose = FALSE) %>%
  tax_transform(rank = "family", trans = "clr", zero_replace = "halfmin") %>%
  ord_calc(method = "PCA") %>%
  ord_plot(
    alpha = 0.6, size = 3, color = "type", shape = "ssx_resolved",
    plot_taxa = 1:4, tax_vec_length = 0.3,
    tax_lab_style = tax_lab_style(
      type = "text", max_angle = 90, aspect_ratio = 0.7,
      size = 3, fontface = "bold"
    ),
  ) +
  theme_classic(12) +
  coord_fixed(0.7, clip = "off")
pca
```

How to interpret the taxa loading vectors?
Cautiously.

There are caveats and nuance to the interpretation of these plots, which are called PCA bi-plots, and you can read more about here: <https://sites.google.com/site/mb3gustame/indirect-gradient-analysis/pca>

The relative length and direction of an arrow indicates how much that taxon contributes to the variation on each visible PC axis, e.g.
Variation in Lacnospiracae contributes to variation along the PC1 axis.

This allows you to infer that samples positioned at the bottom of the plot will tend to have higher relative abundance of Enterococcus than samples at the top of the plot.

Interestingly, samples on the right of the plot (which tend to be donors) seem to have relatively more Lacnospiracae, whilst the recipients have relatively more Morganellacae, Enterobacteracae, and Acidaminococcacae. 

In past studies, "the gut microbiota of people suffering from CDI showed reductions in the abundances of Lachnospiraceae, Ruminococcaceae, Blautia spp., Prevotella spp., Dialister spp., Bifidobacterium spp., Roseburia spp., Anaerostipes spp., Faecalibacterium spp. and Coprococcus spp., in comparison with healthy people. Furthermore, increases in the abundances of Enterococcaceae and Enterococcus were associated with C. difficile infection." doi: 10.3390/pathogens11070781 So, these data make sense.

You might have already noticed this pattern, when exploring and making barplots interactively with `ord_explore` earlier.

### Iris plot

We can make another kind of barplot now, using the PCA information to order our samples in a circular layout.

```{r}
iris <- fmt %>%
  tax_filter(min_prevalence = 2.5 / 100, verbose = FALSE) %>%
  tax_transform(rank = "genus", trans = "clr", zero_replace = "halfmin") %>%
  ord_calc(method = "PCA") %>%
  ord_plot_iris(
    tax_level = "genus", n_taxa = 10, other = "Other",
    anno_colour = "type",
    anno_colour_style = list(alpha = 0.6, size = 1.0, show.legend = FALSE)
  )
iris
```

```{r, fig.height=5, fig.width=10}
patchwork::wrap_plots(pca, iris, nrow = 1, guides = "collect")
```

# Taxon stats

From the PCA loadings and barplots above, we have some strong suspicions about which taxa have a higher relative abundance in donors than in recipients, and vice versa, but we can also statistically test this.
This is often called "differential abundance" testing, in the style of "differential expression" testing from the transcriptomics field.

```{r}
fmt %>%
  comp_barplot(
    tax_level = "genus", n_taxa = 10, facet_by = "type",
    label = NULL, bar_outline_colour = NA
  ) +
  coord_flip() +
  theme(axis.ticks.y = element_blank())
```

## Model one taxon

We will start by creating a linear regression model for one genus, Bacteroides.
We will transform the count data by first making it proportions, and then taking the binary logarithm, log2, after adding a pseudocount.

```{r}
bacteroidesRegression1 <- fmt %>%
  tax_transform("compositional", rank = "genus") %>%
  tax_model(
    type = "lm", rank = "genus",
    trans = "log2", trans_args = list(zero_replace = "halfmin"),
    taxa = "Bacteroides", variables = "type",
    return_psx = FALSE
  ) %>%
  pluck(1)

# looking at the regression results
summary(bacteroidesRegression1)
confint(bacteroidesRegression1)
broom::tidy(bacteroidesRegression1, conf.int = TRUE)
```

<details>

<summary>

**Click here for optional `ggplot2` extension exercise:**

</summary>

Starting from a dataframe like the one produced by the code below, plot:

1.  Easy: The percentage prevalence of Bacteroides in each donor/recipient group
2.  Medium: The distribution of relative abundance of Bacteroides in each donor/recipient group, omitting zeros, on a log2 scale
3.  Hard: Do task 1 or 2 for for several taxa in one plot - (hint: `pivot_longer`)

``` r
fmt %>%
  tax_transform("compositional", rank = "genus") %>%
  ps_get() %>%
  ps_otu2samdat(taxa = "Bacteroides") %>%
  samdat_tbl()
```

</details>

We can fit a model with covariates, as we did for PERMANOVA.
We are going to convert the categorical variables into indicator (dummy) variables, and scale the continuous covariates to 0 mean and SD 1 (z-scores).
You'll see this will make our subsequent plots easier to interpret later.

```{r}
fmt <- fmt %>%
  ps_mutate(
    Recipient = if_else(type == "recipient", true = 1, false = 0),
    Resolved = if_else(ssx_resolved == "Y", true = 1, false = 0)
  )
```

```{r}
bacteroidesRegression2 <- fmt %>%
  tax_transform("compositional", rank = "genus") %>%
  tax_model(
    type = "lm", rank = "genus",
    trans = "log2", trans_args = list(zero_replace = "halfmin"),
    taxa = "Bacteroides", variables = c("type", "ssx_resolved"),
    return_psx = FALSE
  ) %>%
  pluck(1)

# looking at the regression results
summary(bacteroidesRegression2)
confint(bacteroidesRegression2)
broom::tidy(bacteroidesRegression2, conf.int = TRUE)
```


Copy your notebook into the current working directory. Be sure to change to your NETID.

Also, take a minute to "knit" the Rmd notebook into word, and submit this on D2L. Note that this will become useful later for your report, when you select a few graphs to include.

```{bash, engine.opts='-l'}
# copy your notebook into the working directory
cp ~/hw19_pca_taxon_stats.Rmd /xdisk/bhurwitz/bh_class/YOUR_NETID/assignments/19_pca_taxon_stats
```

# Session info

Records your package versions etc.
Useful for debugging / reproducing analysis.

```{r}
devtools::session_info()
```
