---
title: "HW15: beta diversity 1"
author: "Bonnie Hurwitz"
date: "`r Sys.Date()`"
output:
  word_document:
    toc: yes
    toc_depth: '3'
  pdf_document:
    toc: yes
    toc_depth: '3'
  html_document:
    theme: paper
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: yes
subtitle: Exploring Beta Diversity Part 1
editor_options:
  markdown:
    wrap: sentence
---

# Setup

```{r setup, include=FALSE}
# Change your NETID
knitr::opts_chunk$set(echo = TRUE, fig.width = 6, fig.height = 4, dpi = 120)
knitr::opts_knit$set(root.dir = '/xdisk/bhurwitz/bh_class/NETID/assignments/15_beta_diversity_1')
```

```{r}
getwd()
```

```{r message=FALSE}
library(seriation)
library(dplyr)
library(purrr)
library(ggplot2)
library(phyloseq)
library(microViz)
library(shiny)
```

## Importing and fixing taxa your FMT dataset

To get started, we need to import our example.biom file into R and work with it using the phyloseq package

```{r}
# replace your NETID and SETID here, for example set21.biom
fmt <- import_biom("/xdisk/bhurwitz/bh_class/NETID/assignments/12_phyloseq/SETID.biom")
```

### Fixing tax_table

Let's format the dataset.

```{r}
fmt@tax_table@.Data <- substring(fmt@tax_table@.Data, 4)
colnames(fmt@tax_table@.Data)<- c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species")
fmt <- tax_fix(fmt, verbose = FALSE)
```

### Starting here

# Beta diversity (part 1)

β diversity is the difference (measured as distance) between two or more environments. 

It can be measured with metrics like Bray-Curtis dissimilarity, Jaccard distance, or UniFrac distance, to name a few. Each one 
of this measures are focused on a characteristic of the community (*e.g.,* Unifrac distance measures the phylogenetic relationship
between the species of the community).

## Filtering

First, we need to talk about filtering.

### Sample filtering

First, we need to check if any of your samples have a surprisingly low total number of (classified) reads.
This can suggest that something went wrong in the lab (or during sample collection) and the data from this sample might be unreliable.

```{r, fig.height=3, fig.width=6}
fmt %>% 
  ps_mutate(reads = sample_sums(fmt)) %>% 
  samdat_tbl() %>% 
  ggplot(aes(x = reads)) + 
  geom_freqpoly(bins = 500) +
  geom_rug(alpha = 0.5) +
  scale_x_log10(labels = scales::label_number()) +
  labs(x = 'Number of classified reads', y = NULL) +
  theme_bw()
```

How many is enough?
There is no easy answer.

These samples have great depth. There are a few with much less reads than the rest, and a few with under a million.
You might consider dropping the samples with under a million reads, to see if it affects your results, but in this case we won't.

### Taxon filtering

Next, we want to filter out **rare** taxa, before performing our analysis.

#### Why remove rare taxa?

**Rare taxa might be:**

1.  Rare taxa might be erroneous: sequencing errors, chimeric sequences...

2.  Statistical challenges:

    -   There are often A LOT of different rare taxa
        -   Sparsity, lots of zeros
        -   Low precision in very low abundance taxa (integer counts)
        -   Multiple comparisons (we'll get to that later)
    -   Having many many taxa in your dataset can really slow down some analyses!

3.  Biological (ir)relevance:

    -   For many (most?) microbiome research questions, rare taxa are less likely to be relevant
    -   For common host health problems and the human/animal microbiome, if a taxon only occurs in every 100th person, it is less likely to be relevant to risk common disease outcomes (?)


#### How to remove rare taxa?

**What is rare?** Two main concepts.

-   Low **prevalence** - taxon only detected in a small number of samples in your dataset.
-   Low **abundance** - relatively few reads assigned to that taxon (on average or in total)

Consider the impact of issues 1, 2, and 3. Let's say we are not interested in unique taxa that occur in fewer than 2% of samples, and they have to have at least 10,000 reads in total across all samples.

```{r}
# before filtering
ntaxa(fmt)

# after filtering
fmt %>% 
  tax_filter(min_prevalence = 5 / 100, min_total_abundance = 10000) %>% 
  ntaxa()
```

Wow so that would remove **most** of our unique taxa!
What is going on? Let's make some plots!

```{r}
# make table of summary statistics for the unique taxa in fmt
# Nothing will be output, this is used in the next step
fmtTaxaStats <- tibble(
  taxon = taxa_names(fmt),
  prevalence = microbiome::prevalence(fmt),
  total_abundance = taxa_sums(fmt)
)
```

```{r}
p <- fmtTaxaStats %>%
  ggplot(aes(total_abundance, prevalence)) +
  geom_point(alpha = 0.5) +
  geom_rug(alpha = 0.1) +
  scale_x_continuous(
    labels = scales::label_number(), name = "Total Abundance"
  ) +
  scale_y_continuous(
    labels = scales::label_percent(), breaks = scales::breaks_pretty(n = 9),
    name = "Prevalence (%)",
    sec.axis = sec_axis(
      trans = ~ . * nsamples(fmt), breaks = scales::breaks_pretty(n = 9),
      name = "Prevalence (N samples)"
    )
  ) +
  theme_bw()
p
```

So most taxa have a low prevalence, and handful have way more reads than most.

Let's label those points to check which taxa are the big time players.

```{r}
p + ggrepel::geom_text_repel(
  data = function(df) filter(df, total_abundance > 5e6 | prevalence > 1.0),
  mapping = aes(label = taxon), size = 2.5, min.segment.length = 0, force = 15
)
```

Now let's zoom in on the less abundant taxa by log-transforming the axes.
We'll also add lines indicating the thresholds of 5% prevalence and 10000 reads abundance.

```{r}
fmtTaxaStats %>%
  ggplot(aes(x = total_abundance, y = prevalence)) +
  geom_vline(xintercept = 10000, color = "red", linetype = "dotted") +
  geom_hline(yintercept = 5 / 100, color = "red", linetype = "dotted") +
  geom_point(alpha = 0.5) +
  geom_rug(alpha = 0.1) +
  scale_x_log10(labels = scales::label_number(), name = "Total Abundance") +
  scale_y_log10(
    labels = scales::label_percent(), breaks = scales::breaks_log(n = 9),
    name = "Prevalence (%)",
    sec.axis = sec_axis(
      trans = ~ . * nsamples(fmt), breaks = scales::breaks_log(n = 9),
      name = "Prevalence (N samples)"
    )
  ) +
  theme_bw()
```

We can break this down by phylum if we add the taxonomic table information.

```{r, fig.height = 5, fig.width=8}
# don't worry about this code if it's confusing, just focus on the plot output
fmt %>%
  tax_table() %>%
  as.data.frame() %>%
  as_tibble(rownames = "taxon") %>%
  left_join(fmtTaxaStats, by = "taxon") %>%
  add_count(Phylum, name = "phylum_count", sort = TRUE) %>%
  mutate(Phylum = factor(Phylum, levels = unique(Phylum))) %>% # to fix facet order
  mutate(Phylum = forcats::fct_lump_n(Phylum, n = 5)) %>%
  mutate(Phylum = forcats::fct_explicit_na(Phylum, na_level = "Other")) %>%
  ggplot(aes(total_abundance, prevalence)) +
  geom_vline(xintercept = 10000, color = "red", linetype = "dotted") +
  geom_hline(yintercept = 5 / 100, color = "red", linetype = "dotted") +
  geom_point(alpha = 0.5, size = 1) +
  geom_rug(alpha = 0.2) +
  scale_x_log10(
    labels = scales::label_log(), breaks = scales::breaks_log(n = 5),
    name = "Total Abundance"
  ) +
  scale_y_log10(
    labels = scales::label_percent(), breaks = scales::breaks_log(n = 9),
    name = "Prevalence (%)",
    sec.axis = sec_axis(
      trans = ~ . * nsamples(fmt), breaks = scales::breaks_log(n = 9),
      name = "Prevalence (N samples)"
    )
  ) +
  facet_wrap("Phylum") +
  theme_bw(10)
```

**How to pick a threshold?**

Depends on what analysis method you are filtering for!

-   alpha diversity = DO NOT FILTER
-   beta diversity = relevance of threshold depends on your distance measure
-   differential abundance testing = stringent filtering, prevalence \>5%, \>10%?

## Finishing up

Copy your notebook into the current working directory. Be sure to change to your NETID.

Also, take a minute to "knit" the Rmd notebook into word, and submit this on D2L. Note that this will become useful later for your report, when you select a few graphs to include.

```{bash, engine.opts='-l'}
# copy your notebook into the working directory
cp ~/hw15_beta_diversity_1.Rmd /xdisk/bhurwitz/bh_class/NETID/assignments/beta_diversity_1
```

# Session info

Records your package versions etc.
Useful for debugging / reproducing analysis.

```{r}
devtools::session_info()
```
