---
title: "HW16: beta diversity 2"
author: "Bonnie Hurwitz"
date: "`r Sys.Date()`"
output:
  word_document:
    toc: yes
    toc_depth: '3'
  pdf_document:
    toc: yes
    toc_depth: '3'
  html_document:
    theme: paper
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: yes
subtitle: Exploring Beta Diversity Part 2
editor_options:
  markdown:
    wrap: sentence
---

# Setup

```{r setup, include=FALSE}
# Change your NETID
knitr::opts_chunk$set(echo = TRUE, fig.width = 6, fig.height = 4, dpi = 120)
knitr::opts_knit$set(root.dir = '/xdisk/bhurwitz/bh_class/NETID/assignments/16_beta_diversity_2')
```

```{r}
getwd()
```

```{r message=FALSE}
library(seriation)
library(dplyr)
library(purrr)
library(ggplot2)
library(phyloseq)
library(microViz)
library(shiny)
```

## Importing and fixing taxa your FMT dataset

To get started, we need to import our example.biom file into R and work with it using the phyloseq package

```{r}
# replace your NETID and SETID here, for example set21.biom
fmt <- import_biom("/xdisk/bhurwitz/bh_class/NETID/assignments/12_phyloseq/SETID.biom")
```

### Fixing tax_table

Let's format the dataset.

```{r}
fmt@tax_table@.Data <- substring(fmt@tax_table@.Data, 4)
colnames(fmt@tax_table@.Data)<- c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species")
fmt <- tax_fix(fmt, verbose = FALSE)
```

# Beta diversity (part 2)

## Dissimilarity measures

What are we doing? Calculating the dissimilarity of two samples' compositions (pairwise distance).

Diversity β measures how different two or more communities are, either in their composition (richness) 
or in the abundance of the organisms that compose it (abundance). 

There are several dissimilarity measures we can use:

-   Binary Jaccard - presence-absence only
-   Bray-Curtis - abundance weighted
-   UniFrac distances (unweighted, weighted, generalised)

More details on each distance metric:

- Jaccard distance: Based on the presence/absence of species (diversity). 
It goes from 0 (same species in the community) to 1 (no species in common)

- Bray-Curtis dissimilarity: The difference in richness and abundance across environments (samples). Weight on abundance. 
Measures the differences from 0 (equal communities) to 1 (different communities)

- UniFrac: Measures the phylogenetic distance; how alike the trees in each community are. 
There are two types, without weights (diversity) and with weights (diversity and abundance)  

Note that we are going to filter out rare taxa for each analysis rather than filter the entire fmt dataset.

### Jaccard distances

Let's start out by calcualting the Jaccard Distances.

```{r}
fmt %>%
  tax_filter(min_prevalence = 2.5 / 100) %>%
  tax_agg(rank = "Genus") %>%
  tax_transform("binary") %>% # converts counts to absence/presence: 0/1
  dist_calc(dist = "jaccard")
```

So we get our pairwise dissimilarities, in a distance matrix 

```{r}
jdistances <- fmt %>%
  tax_filter(min_prevalence = 2.5 / 100, verbose = FALSE) %>%
  tax_agg(rank = "Genus") %>%
  tax_transform("binary") %>%
  dist_calc(dist = "jaccard") %>%
  dist_get()
```

You can extract the dissimilarities or distances with dist_get. Change the matrix size based on the number of samples you have.

```{r}
sample_names(fmt) # count the number of sample names
```

```{r}
as.matrix(jdistances)[1:10, 1:10]
```

The Binary Jaccard dissimilarities range between zero (identical) and one (no shared genera). Let's see what the range is for our pairwise sample distances.

```{r}
range(jdistances)
```


### Bray-Curtis Distances

Next, let's account for abundance and calculate the Bray-Curtis Distances.

```{r}
fmt %>%
  tax_filter(min_prevalence = 2.5 / 100) %>%
  tax_agg(rank = "Genus") %>%
  dist_calc(dist = "bray")
```

Hiw do the distances change when we account for abundance? Let's create the distance matrix. 


```{r}
bdistances <- fmt %>%
  tax_filter(min_prevalence = 2.5 / 100, verbose = FALSE) %>%
  tax_agg(rank = "Genus") %>%
  dist_calc(dist = "bray") %>%
  dist_get()
```

You can extract the dissimilarities or distances with dist_get. Change this based on the number of samples you have.

```{r}
as.matrix(bdistances)[1:10, 1:10]
```

What is the range of values that we see in the pairwise distance matrix? Why is this different from the Jaccard distance?

```{r}
range(bdistances)
```

## Finishing up

Copy your notebook into the current working directory. Be sure to change to your NETID.

Also, take a minute to "knit" the Rmd notebook into word, and submit this on D2L. Note that this will become useful later for your report, when you select a few graphs to include.

```{bash, engine.opts='-l'}
# copy your notebook into the working directory
cp ~/hw16_beta_diversity_2.Rmd /xdisk/bhurwitz/bh_class/NETID/assignments/beta_diversity_2
```

# Session info

Records your package versions etc.
Useful for debugging / reproducing analysis.

```{r}
devtools::session_info()
```
